---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

As this is a package extension, if you are new to `{admiral}` then the best
place to first start reading would be the [Get
Started](https://pharmaverse.github.io/admiral/articles/admiral.html) page. This
extension package follows the same main idea and conventions, and re-uses many
functions from `{admiral}`, so it is important to thoroughly understand these to
be able to use `{admiralneuro}`.

# Creating Neuroscience ADaM Datasets

For the neuroscience ADaM data structures, an overview of the flow and example function calls
for the most common steps are provided by the following vignette:

- [Creating ADAPET/ADTPET](adpet.html)


`{admiralneuro}` also provides template R scripts as a starting point. They can be
created by calling `use_ad_template()` from `{admiral}`, e.g.,


```{r, eval = FALSE}
library(admiral)

use_ad_template(
  adam_name = "adapet",
  save_path = "./ad_adapet.R",
  package = "admiralneuro"
)
```

A list of all available templates can be obtained by `list_all_templates()`
from `{admiral}`:
  
```{r, eval = FALSE}
list_all_templates(package = "admiralneuro")
```

# Support

Support is provided via the [admiral Slack channel](https://pharmaverse.slack.com/). Additionally, please feel free to raise issues in our [GitHub repository](https://github.com/pharmaverse/admiralneuro/issues).
