---
title: "Creating ADNV"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating ADNV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This article describes the creation of the `ADNV` (Nervous System Analysis Dataset) ADaM.

The `ADNV` dataset focuses on olfactory function assessments, specifically the University of Pennsylvania Smell Identification Test (UPSIT). This vignette demonstrates the neuro-specific steps for creating this BDS finding dataset.

We advise you first consult the `{admiral}` [Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html). The programming workflow around creating the general set-up of a BDS ADaM dataset using `{admiral}` functions is the same. In this vignette, we focus on the neuro-specific steps to avoid repeating information and maintaining the same content in two places. As such, the code in this vignette is not completely executable; we recommend consulting the `ADNV` template script to view the full workflow.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless otherwise specified.*

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(admiralneuro)
library(pharmaversesdtm)
library(dplyr)
library(lubridate)
library(stringr)
```

# Programming Workflow

-   [Read in Data](#readdata)
-   [Initial Setup](#adnv_setup)
-   [Assign `PARAMCD`, `PARAM`, and `PARAMN`](#paramcd)
-   [Map `AVAL` and `AVALC` variables](#aval)
-   [Convert UPSIT Score to Percentile](#percentile)
-   [Derive Criterion Flag Variables for UPSIT Percentile](#criterionfl)
-   [Remaining Analysis Dataset Set-up](#dataset_end)

## Read in Data {#readdata}

To start, all data frames needed for the creation of `ADNV` should be loaded into the global environment. Reading data will usually be a company specific process, however, for the purpose of this vignette, we will use example data from `{pharmaversesdtm}` and `{admiralneuro}`. We will utilize `NV` and `ADSL` for the basis of the Nervous System Analysis dataset.

```{r, message=FALSE, warning=FALSE}
nv <- pharmaversesdtm::nv_neuro %>%
  convert_blanks_to_na()

adsl <- admiralneuro::adsl_neuro %>%
  convert_blanks_to_na()
```

## Initial Setup {#adnv_setup}

The following steps are used to merge `ADSL` variables with the source data `NV`,  followed with filtering to retain only UPSIT records. Common analysis variables such as Analysis Date (`ADT`) and Relative Analysis Day (`ADY`) can also be derived during this process. Note that only the sections relevant to this vignette are included in the steps below. To get detailed guidance on all the steps, refer to the `{admiral}` [Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).



```{r, message=FALSE, warning=FALSE}
adsl_vars <- exprs(AGE, SEX, TRTSDT, TRTEDT, TRT01A, TRT01P)

adnv <- nv %>%
  derive_vars_merged(
    dataset_add = adsl,
    new_vars = adsl_vars,
    by_vars = exprs(STUDYID, USUBJID)
  ) %>%
  filter(NVTESTCD == "UPSIT") %>%
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc = NVDTC
  ) %>%
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT))
```

### Assign `PARAMCD`, `PARAM`, and `PARAMN` variables {#paramcd}

The next step is to create and assign parameter level variables such as `PARAMCD`, `PARAM`, and `PARAMN`. For this, a lookup can be created based on the SDTM `--TESTCD` and `--CAT` values to join to the source data.

```{r, echo=TRUE, message=FALSE}
param_lookup <- tibble::tribble(
  ~NVTESTCD, ~NVCAT, ~PARAMCD, ~PARAM, ~PARAMN,
  "UPSIT", "OLFACTORY FUNCTION", "UPSITTS", "UPSIT Combined Score from 40 Odorant", 1
)
```

This lookup may now be joined to the data:

```{r, eval=TRUE, include=TRUE, message=FALSE}
adnv <- adnv %>%
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    new_vars = exprs(PARAMCD, PARAM, PARAMN),
    by_vars = exprs(NVTESTCD, NVCAT)
  )
```

```{r, eval=TRUE, echo=FALSE}
adnv_param <- distinct(adnv, USUBJID, NVTESTCD, NVCAT, PARAMCD, PARAM, PARAMN)
dataset_vignette(adnv_param, display_vars = exprs(USUBJID, NVTESTCD, NVCAT, PARAMCD, PARAM, PARAMN))
```

### Map `AVAL` and `AVALC` variables {#aval}

Now that the parameter-level variables have been created, we can map the Analysis Value variables. For these parameters, no complex derivation is required. Note that `AVALC` should only be mapped if it contains non-redundant information.

```{r, message=FALSE, warning=FALSE}
adnv <- adnv %>%
  mutate(
    AVAL = NVSTRESN,
    AVALC = if_else(
      is.na(NVSTRESN) | as.character(NVSTRESN) != NVSTRESC,
      NVSTRESC,
      NA
    )
  )
```

We then drop AVALC column because it contains no non-missing values

```{r, message=FALSE, warning=FALSE}
if (all(is.na(adnv$AVALC))) {
  adnv <- adnv %>% select(-AVALC)
}
```

## Convert UPSIT Score to Percentile {#percentile}

The **UPSIT (University of Pennsylvania Smell Identification Test)** is a standardized test that measures olfactory function using 40 odor identification items. The raw UPSIT score ranges from 0 to 40, with higher scores indicating better olfactory function.

To enable meaningful interpretation across different age and sex groups, raw UPSIT scores can be converted to **percentiles** based on normative data. These percentiles represent where an individual's performance ranks relative to age- and sex-matched peers.

In the following steps, we will incorporate the **conversion** from **UPSIT total score to percentile** using the function `admiralneuro::compute_upsit_percentile()`.

**1. Define the Source Variables**

The first step is to define `keep_vars`, which lists the source variables to retain and copy into newly derived records in the next step; all other variables will remain in the dataset but will be set to `NA` in those derived rows.

**2. Convert UPSIT Score to Percentile**

The second step creates new analysis records in which the original `UPSITTS` values are converted into percentiles using `admiralneuro::compute_upsit_percentile()`. The conversion is based on the patient's sex and age at the time of assessment.

The new parameter values which include `AVAL`, `PARAMCD`, `PARAM`, and `PARAMN` are assigned for the percentile parameter.

```{r, message=FALSE, warning=FALSE}
keep_vars <- c(
  get_admiral_option("subject_keys"),
  adsl_vars,
  exprs(ADT, ADY, VISIT)
)

# Prepare derived rows
upsit_percentile <- adnv %>%
  filter(PARAMCD == "UPSITTS") %>%
  mutate(
    AVAL = compute_upsit_percentile(
      sex = SEX,
      age = AGE,
      upsit_score = AVAL
    ),
    PARAMCD = "UPSITPC",
    PARAM = "Percentile derived from UPSIT total score",
    PARAMN = 2,
    AVALU = NA
  )

# Bind with original dataset
adnv <- bind_rows(adnv, upsit_percentile)

# Drop AVALU because it contains no non-missing values
if (all(is.na(adnv$AVALU))) {
  adnv <- adnv %>% select(-AVALU)
}
```

This is how the parameters and newly converted parameters will look like:
```{r, eval=TRUE, echo=FALSE}
adnv_param <- distinct(adnv, USUBJID, PARAMCD, PARAM, AVAL, ADT, ADY, VISIT)
dataset_vignette(adnv_param %>% arrange(USUBJID, ADT), display_vars = exprs(USUBJID, PARAMCD, PARAM, AVAL, ADT, ADY, VISIT))
```

### Derive Criterion Flag Variables for UPSIT Percentile {#criterionfl}

Having converted the UPSIT score to percentile, we can now derive the criterion flags for olfactory impairment categories.

We will use a percentile cutoff of **10** to define olfactory impairment. The 10th percentile cutoff represents a sex- and age-adjusted threshold for olfactory impairment based on normative data. UPSIT percentile ≤ 10 is considered olfactory impairment, while UPSIT percentile > 10 is considered normal olfactory function.

To do this, we will use the `admiral::derive_vars_crit_flag()` function. Since we want to derive these flags specifically for the UPSIT percentile parameter, we will apply the derivation only to those records using `admiral::restrict_derivation()`.

```{r, message=FALSE, warning=FALSE}
adnv <- adnv %>%
  restrict_derivation(
    derivation = derive_vars_crit_flag,
    args = params(
      crit_nr = 1,
      condition = if_else(PARAMCD == "UPSITPC", AVAL <= 10, NA),
      description = "UPSITPC <= 10",
      values_yn = TRUE # To get "Y", "N", and NA for the flag
    ),
    filter = PARAMCD == "UPSITPC"
  )
```

This is how the criterion flags will look like:

```{r, eval=TRUE, echo=FALSE}
adnv_flags <- distinct(adnv, USUBJID, PARAMCD, PARAM, AVAL, ADT, ADY, VISIT, CRIT1, CRIT1FL)
dataset_vignette(adnv_flags %>% arrange(USUBJID, ADT), display_vars = exprs(USUBJID, PARAMCD, PARAM, AVAL, ADT, ADY, VISIT, CRIT1, CRIT1FL))
```

## Remaining Analysis Dataset Set-up {#dataset_end}

The `{admiral}` [Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html) covers all the steps that are not shown here, such as merging the timing variables, analysis flags, baseline derivations, change from baseline calculations, etc.

# Example Scripts

| ADaM   | Sourcing Command                                               |
|--------|----------------------------------------------------------------|
| ADNV | `admiral::use_ad_template("ADNV", package = "admiralneuro

## References

University of Pennsylvania Smell Identification Test. (n.d.). In *Wikipedia*. Retrieved from https://en.wikipedia.org/wiki/University_of_Pennsylvania_Smell_Identification_Test

Brumm, M. C. *et al.* (2023). Updated Percentiles for the University of Pennsylvania Smell Identification Test in Adults 50 Years of Age and Older. *Neurology, 100*(16), e1691–e1701.

Mastenbroek, S. E. *et al.* (2025). Two-step detection of Lewy body pathology via smell-function testing and CSF α-synuclein seed amplification. *Nature Communications, 16*, 7182.
