---
title: "Creating ADTPET and ADAPET"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating ADTPET and ADAPET}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This article describes the creation of PET Scan ADaMs.

We recommend splitting out Amyloid and Tau into two distinct ADaMs datasets. The vignette presents
a combined approach for the initial setup, parameter mapping and finals steps.
It then splits into separate sections for each data type, as they require different processes:
starting with `ADTPET` (Tau PET Scan Analysis Dataset), followed by `ADAPET` (Amyloid PET Scan Analysis Dataset).

We advise you first consult the `{admiral}` [Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).
The programming workflow around creating the general set-up of a BDS ADaM dataset using
`{admiral}` functions is the same. In this vignette, we focus on the neuro-specific steps only to avoid repeating
information and maintaining the same content in two places. As such, the code in this vignette is not
completely executable; we recommend consulting the ADTPET and ADAPET template scripts
to view the full workflow.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input
unless otherwise specified.*

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(admiralneuro)
library(pharmaversesdtm)
library(dplyr)
library(lubridate)
library(stringr)
library(metatools)
```

# Programming Workflow

-   [Read in Data](#readdata)
-   [Initial PET Scan Analysis Dataset Set-up](#adpet_setup)
-   [Assign `PARAMCD`, `PARAM`, and `PARAMN`](#paramcd)
-   [Map `AVAL` and `AVALC` variables](#aval)
-   [`ADTPET` (Tau PET Scan Analysis Dataset)](#adtpet)
-   [`ADAPET` (Amyloid PET Scan Analysis Dataset)](#adapet)
    -   [Convert SUVR to Centiloid](#centiloid)
    -   [Derive Criterion Flag Variables for Amyloid Categories](#criterionfl)
-   [Remaining PET Scan Analysis Dataset Set-up](#dataset_end)

## Read in Data {#readdata}

To start, all data frames needed for the creation of `ADTPET` or `ADAPET` should be
loaded into the global environment. Reading data will usually be a
company specific process, however, for the purpose of this vignette, we
will use example data from `{admiralneuro}`. We will
utilize `NV`, `SUPPNV`, `AG` and `ADSL` for the basis of the Pet Scan ADaM datasets.

```{r, message=FALSE, warning=FALSE}
nv <- admiralneuro::nv_neuro %>%
  convert_blanks_to_na()

suppnv <- admiralneuro::suppnv_neuro %>%
  convert_blanks_to_na()

ag <- admiralneuro::ag_neuro %>%
  convert_blanks_to_na()

adsl <- admiralneuro::adsl_neuro %>%
  convert_blanks_to_na()
```

While loading our input data, we can combine the `NV` domain and the `SUPPNV` supplementary domain for easier use in the next steps. Using the `metatools::combine_supp()` function avoids the need to manually transpose and merge the supplementary dataset with the corresponding domain.

```{r, message=FALSE, warning=FALSE}
nv_suppnv <- combine_supp(nv, suppnv)
```

## Initial PET Scan Analysis Dataset Set-up {#adpet_setup}

The following steps are used to merge `ADSL` variables with the source data `NV` (previously combined with `SUPPNV`),
along with tracer information from the `AG` dataset, without distinguishing between
Tau and Amyloid data at this stage. This distinction will be made later.
Common analysis variables such as Analysis Date (`ADT`) and Relative Analysis Day (`ADY`)
can also be derived during this process.
Note that only the sections relevant to this vignette are included in the steps below.
To get a detailed guidance on all the steps, refer the `{admiral}`
[Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).

```{r, message=FALSE, warning=FALSE}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

adpet <- nv_suppnv %>%
  derive_vars_merged(
    dataset_add = adsl,
    new_vars = adsl_vars,
    by_vars = exprs(STUDYID, USUBJID)
  ) %>%
  derive_vars_merged(
    dataset_add = ag,
    new_vars = exprs(AGTRT, AGCAT),
    by_vars = exprs(STUDYID, USUBJID, VISIT, NVLNKID = AGLNKID)
  ) %>%
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc = NVDTC
  ) %>%
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT))
```

### Create `PARAMCD`, `PARAM`, and `PARAMN` variables {#paramcd}

The next step is to create and assign parameter level variables such as
`PARAMCD`, `PARAM`, and `PARAMN`. For this, a lookup can be
created based on the SDTM `--TESTCD`, `--CAT`, `--LOC`, `--METHOD` and `REFREG` values to join to the source data.

```{r, echo=TRUE, message=FALSE}
param_lookup <- tibble::tribble(
  ~NVTESTCD, ~NVCAT, ~NVLOC, ~REFREG, ~NVMETHOD, ~PARAMCD, ~PARAM, ~PARAMN,
  "SUVR", "FBP", "NEOCORTICAL COMPOSITE", "Whole Cerebellum", "AVID FBP SUVR PIPELINE", "SUVRAFBP", "AVID FBP Standard Uptake Ratio Neocortical Composite Whole Cerebellum", 1,
  "SUVR", "FBB", "NEOCORTICAL COMPOSITE", "Whole Cerebellum", "AVID FBB SUVR PIPELINE", "SUVRAFBB", "AVID FBB Standard Uptake Ratio Neocortical Composite Whole Cerebellum", 2,
  "SUVR", "FBP", "NEOCORTICAL COMPOSITE", "Whole Cerebellum", "BERKELEY FBP SUVR PIPELINE", "SUVRBFBP", "BERKELEY FBP Standard Uptake Ratio Neocortical Composite Whole Cerebellum", 3,
  "SUVR", "FBB", "NEOCORTICAL COMPOSITE", "Whole Cerebellum", "BERKELEY FBB SUVR PIPELINE", "SUVRBFBB", "BERKELEY FBB Standard Uptake Ratio Neocortical Composite Whole Cerebellum", 4,
  "SUVR", "FTP", "NEOCORTICAL COMPOSITE", "Inferior Cerebellar Gray Matter", "BERKELEY FTP SUVR PIPELINE", "SUVRBFTP", "BERKELEY FTP Standard Uptake Ratio Neocortical Composite Inferior Cerebellar Gray Matter", 5,
  "VR", "FBP", NA, NA, "FBP VISUAL CLASSIFICATION", "VRFBP", "FBP Qualitative Visual Classification", 6,
  "VR", "FBB", NA, NA, "FBB VISUAL CLASSIFICATION", "VRFBB", "FBB Qualitative Visual Classification", 7,
  "VR", "FTP", NA, NA, "FTP VISUAL CLASSIFICATION", "VRFTP", "FTP Qualitative Visual Classification", 8
)
```

This lookup may now be joined to the data and this is how the
parameters will look like:

```{r, eval=TRUE, include=TRUE, message=FALSE}
adpet <- adpet %>%
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    new_vars = exprs(PARAMCD, PARAM, PARAMN),
    by_vars = exprs(NVTESTCD, NVCAT, NVLOC, NVMETHOD, REFREG)
  )
```

```{r, eval=TRUE, echo=FALSE}
adpet_param <- distinct(adpet, USUBJID, NVTESTCD, NVCAT, NVLOC, NVMETHOD, REFREG, PARAMCD, PARAM, PARAMN)
dataset_vignette(adpet_param, display_vars = exprs(USUBJID, NVTESTCD, NVCAT, NVLOC, NVMETHOD, REFREG, PARAMCD, PARAM, PARAMN))
```

### Map `AVAL` and `AVALC` variables {#aval_tpet}

Now that the parameter-level variables have been created, we can map the Analysis Value variables.
For these parameters, no complex derivation is required.
Note that `AVALC` should only be mapped if it contains non-redundant information.

```{r, message=FALSE, warning=FALSE}
adpet <- adpet %>%
  mutate(
    AVAL = NVSTRESN,
    AVALC = if_else(
      is.na(NVSTRESN) | as.character(NVSTRESN) != NVSTRESC,
      NVSTRESC,
      NA
    )
  )
```

Having established the foundation of a global PET Scan Analysis Dataset (`adpet`),
we can now differentiate between Tau and Amyloid data.

## `ADTPET` (Tau PET Scan Analysis Dataset){#adtpet}

Tau PET scan data generally do not require any further derivations.
Therefore, if producing an `ADTPET`, we can now filter the global PET Scan Analysis Dataset `adpet` to retain only the Tau tracer records.

```{r, message=FALSE, warning=FALSE}
adtpet <- adpet %>%
  filter(AGCAT == "TAU TRACER")
```

## `ADAPET` (Amyloid PET Scan Analysis Dataset){#adapet}

If producing an `ADAPET`, we can now filter the global PET Scan Analysis Dataset `adpet` to retain only the Amyloid tracer records.

```{r, message=FALSE, warning=FALSE}
adapet <- adpet %>%
  filter(AGCAT == "AMYLOID TRACER")
```

### Convert SUVR to Centiloid{#centiloid}

`SUVR` (Standardized Uptake Value Ratio) is a widely used semi-quantitative measure
in PET imaging that estimates the level of tracer uptake (such as Amyloid) in the brain,
relative to a reference region.
While useful, `SUVR` has certain limitations (which we will not detail here) that affect
its comparability across different tracers and studies.
To address this, the **Centiloid scale** was developed to standardize Amyloid PET measurements,
enabling consistent interpretation across imaging protocols, tracers, and clinical studies.

In the following steps, we will incorporate the **conversion** from **SUVR to Centiloid** using
the function `admiralneuro::compute_centiloid()`.

For our examples, we divide the process into three main steps:

**1. Define the Grouping Criteria**

The first step is to specify how to group the data using `admiral::slice_derivation()`.
In this case, we group by patient, `ADSL` variables, Analysis Date (`ADT`),
Relative Analysis Day (`ADY`), and Visit (`VISIT`).
We also retain records with missing Analysis Values (`AVAL = NA`) by setting `keep_nas = TRUE`.

**2. Convert SUVR to Centiloid: Florbetaben (FBB)**

The second step involves converting the original `SUVR` parameter for Florbetaben
to Centiloid using `admiralneuro::compute_centiloid()`.
We apply filtering conditions using `admiral::derivation_slice()` based on
the tracer (`SUVRBFBB`), method (`BERKELEY FBB SUVR PIPELINE`), and reference region (`Whole Cerebellum`).
The new parameter values which include `AVAL`, `PARAMCD`, `PARAM`, and `AVALU`,
are assigned within the `args` argument of this function.

**3. Convert SUVR to Centiloid: Florbetapir (FBP)**

Similarly, the third step converts the `SUVR` parameter for Florbetapir.
Filtering is applied to select records based on the `SUVRAFBP` tracer, the
`AVID FBP SUVR PIPELINE` method, and the same reference region.
As in step 2, the corresponding Centiloid values and associated metadata are set
within the `args` argument.

```{r, message=FALSE, warning=FALSE}
adapet <- adapet %>%
  slice_derivation(
    derivation = derive_param_computed,
    args = params(
      by_vars = c(
        get_admiral_option("subject_keys"),
        adsl_vars,
        exprs(ADT, ADY, VISIT)
      ),
      keep_nas = TRUE
    ),
    derivation_slice(
      filter = (PARAMCD == "SUVRBFBB") & (NVMETHOD == "BERKELEY FBB SUVR PIPELINE") & (REFREG == "Whole Cerebellum"),
      args = params(
        parameters = c("SUVRBFBB"),
        set_values_to = exprs(
          AVAL = compute_centiloid(
            tracer = "18F-Florbetaben",
            pipeline = "BERKELEY FBB SUVR PIPELINE",
            ref_region = "Whole Cerebellum",
            suvr = AVAL
          ),
          PARAMCD = "CLBFBB",
          PARAM = "Centiloid (CL) based on BERKELEY FBB",
          AVALU = "CL"
        )
      )
    ),
    derivation_slice(
      filter = (PARAMCD == "SUVRAFBP") & (NVMETHOD == "AVID FBP SUVR PIPELINE") & (REFREG == "Whole Cerebellum"),
      args = params(
        parameters = c("SUVRAFBP"),
        set_values_to = exprs(
          AVAL = compute_centiloid(
            tracer = "18F-Florbetapir",
            pipeline = "AVID FBP SUVR PIPELINE",
            ref_region = "Whole Cerebellum",
            suvr = AVAL
          ),
          PARAMCD = "CLAFBP",
          PARAM = "Centiloid (CL) based on AVID FBP",
          AVALU = "CL"
        )
      )
    )
  )
```

This is how the parameters and newly converted parameters will look like:

```{r, eval=TRUE, echo=FALSE}
adapet_param <- distinct(adapet %>% filter(NVTESTCD == "SUVR" | !is.na(AVALU)), USUBJID, PARAMCD, PARAM, AVAL, AVALU, ADT, ADY, VISIT)
dataset_vignette(adapet_param %>% arrange(USUBJID, ADT), display_vars = exprs(USUBJID, PARAMCD, PARAM, AVAL, AVALU, ADT, ADY, VISIT))
```

### Derive Criterion Flag Variables for Amyloid Categories{#criterionfl}

Having converted the Standardized Uptake Value Ratio to Centiloid, we can now derive
the criterion flags for the Amyloid categories.

To do this, we will use the `admiral::derive_vars_crit_flag()` function.
Since we want to derive these flags specifically for the Centiloid parameters,
we will apply the derivation only to those records using `admiral::restrict_derivation()`.

```{r, message=FALSE, warning=FALSE}
adapet <- adapet %>%
  restrict_derivation(
    derivation = derive_vars_crit_flag,
    args = params(
      crit_nr = 1,
      condition = if_else(PARAMCD %in% c("CLAFBP", "CLBFBB"), AVAL < 24.1, NA),
      description = "CENTILOID < 24.1",
      values_yn = TRUE # To get "Y", "N", and NA for the flag
    ),
    filter = PARAMCD %in% c("CLAFBP", "CLBFBB")
  )
```

This is how the criterion flags will look like:

```{r, eval=TRUE, echo=FALSE}
adapet_flags <- distinct(adapet %>% filter(NVTESTCD == "SUVR" | !is.na(AVALU)), USUBJID, PARAMCD, PARAM, AVAL, AVALU, ADT, ADY, VISIT, CRIT1, CRIT1FL)
dataset_vignette(adapet_flags %>% arrange(USUBJID, ADT), display_vars = exprs(USUBJID, PARAMCD, PARAM, AVAL, AVALU, ADT, ADY, VISIT, CRIT1, CRIT1FL))
```

## Remaining PET Scan Analysis Dataset Set-up{#dataset_end}

The `{admiral}` [Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html)
covers all the steps that are not shown here, such as merging the timing variables,
analysis flags, etc.

# Example Scripts

ADaM | Sourcing Command
---- | --------------
ADTPET | `admiral::use_ad_template("ADTPET", package = "admiralneuro")`
ADAPET | `admiral::use_ad_template("ADAPET", package = "admiralneuro")`
